\Large {\bf Protocol-Guided Analysis of Post-silicon Traces Under Limited Observability}

### 2015-09-30 paper.tex
My pass

\begin{abstract}
We consider the problem of reconstructing system-level
behavior of an SoC design from a partially observed signal
trace.  Solving this problem is a critical activity in
post-silicon validation, and currently depends primarily on
human creativity and insights.  In this paper, we provide an algorithm to
automatically infer system-level transactions from
incomplete, ambiguous, and noisy trace data.
% together with a measure of confidence.  
We demonstrate the approach on a multicore virtual platoform
developed within the GEM5 environment.
\end{abstract}

@@

In this paper, we consider the problem of reconstructing
protocol-level behavior from silicon traces in SoC designs.
Given a collection of system-level communication protocols
and a trace of (partially observed) hardware signals, our
approach infers, with a certain measure of confidence, the
protocol instances (and their interleavings) being exercised
by the trace.  The approach is based on a formalization of
system-level transactions via labeled Petri-Nets, which are
capable of describing sequencing, concurrency, and choices
over system events.  We develop algorithms to infer
system-level transactions from traces with missing, noisy,
and ambiguous signal values.
% together with an estimate of confidence on the inference.  
We demonstrate our approach on a multicore virtual platform
constructed within the GEM5 environment~\cite{Binkert2011}.

@@

An SoC design involves integration of a number of IPs that
communicate through complex protocols.  Such system-level
protocols are typically specified in architecture documents
as message flow diagrams.  As illustration,
Fig~\ref{flow-spec-ex}(a) shows one such diagram for a
protcol to authenticate and load a firmware during system
boot for firmware upgrade.  During validation, the system
under debug (SUD) exercises some complex system-level
use-case which involves interleaved execution of possibly a
large number of such protocols.  A trace of a small number
of hardware signals is then shipped off-chip analysis.  The
off-chip analysis includes two broad phases: (1)~trace
abstraction, and (2)~trace interpretation.  Trace
abstraction maps the hardware trace into higher-level
architectural constructs, \eg, messages, operations, etc.: a
message such as {\tt Authorization request} may be
implemented in hardware through a Boolean or temporal
combination of specific hardware signals in the NoC fabric
between {\tt Device} and {\tt CE}, \eg, as a sequence
containing a header, a specific value of a sequence of data
words, etc.  We will refer to such architectural constructs
as {\em protocol events} or {\em flow events}.  Note that
due to limited observability, it may not be possible to map
a given set of (observed) hardware signals uniquely to a
flow event.  Finally, the trace may a result from several
instances of the same protocol executing concurrently, \eg,
a firmware authentication protocol may be invoked when
another instance of the protocol has not completed.

@@

To illustrate the basic idea of the trace analysis method, 
consider the system flow shown in Figure~Fig.~\ref{flow-spec-ex}(b).
Let $F_1$ denote such flow.  Suppose that a hardware system 
implements flow $F_1$, and the following flow trace is abstracted 
from an observed signal trace as a result of executing such system.
\[
	t_1\;t_2\;t_1\;t_2\;t_3\;t_3\;t_4\;t_5\;t_5\;t_4\ldots
\]  
where transition names in the LPN are used to represent the 
flow events in the trace.  The first four events results in
the following flow execution scenario
\[
	\{(F_{1,1}, \{p_3\}),~(F_{1,2}, \{p_3\})\}.
\]
For the first event $t_3$, it results in two execution scenarios 
below depending on which flow instance emits $t_3$.
\[
\begin{array}{l}
	\{(F_{1,1}, \{p_4\}),~(F_{1,2}, \{p_3\})\} \\
	\{(F_{1,1}, \{p_3\}),~(F_{1,2}, \{p_4\})\}.
\end{array}
\]
After handing the next event $t_3$, the above two execution scenarios
are reduced to the one as shown below.
\[
	\{(F_{1,1}, \{p_4\}),~(F_{1,2}, \{p_4\})\}.
\]
By following Algorithm~\ref{algo:compliance} to handle the remaining 
four events, the following execution scenario can be derived.
\[
	\{(F_{1,1}, \{p_5, p_6\}),~(F_{1,2}, \{p_5, p_6\})\}
\]
%% Given a trace of flow events $\rho = e_1e_2\ldots e_n$, the
%% trace interpretation algorithm starts with an empty set of
%% of flow execution scenario $\mathit{Scen} = \emptyset$.
%% Then, for each $e_h$ where $1 \leq h \leq n$ starting $h=1$,
%% and for each $\mathit{scen} \in \mathit{Scen}$, the
%% following two steps are performed.
%% \begin{description}
%% \item[Step 1]~~For each $(F_{i,j}, s_{i,j}) \in
%%   \mathit{scen}$, if $\mathit{accept}(F_{i,j}, s_{i,j}, e_h)
%%   = (F_{i,j}, s'_{i,j})$, create a new scenario
%%   $\mathit{scen}' = (\mathit{scen} - (F_{i,j}, s_{i,j}))
%%   \cup (F_{i,j}, s'_{i,j})$, which is added into
%%   $\mathit{Scen}'$.

@@

The trace interpretation discussion in the preceding section
assumed full observability.  In this section, we discuss how
to adapt the method to deal with silicon traces under
limited (partial) obsrvability.

@@

Hereafter, the term {\em flow traces} is used to refer to traces of flow events. 
Consider the following example for the first case.  Suppose that there are three flow events: $e_1$, $e_2$, and $e_3$, which are implemented in hardware by the signal events shown in the list below.  We use Boolean expressions to represent signal events for the discussion.
\[
\begin{array}{cl}
e_1: & abc\\
e_2: & \bar{a}bc\\
e_3: & a\bar{b}c
\end{array}
\] 
Now suppose that only signals $b$ and $c$ are observable, and a signal trace of this partial observability is obtained below.
\[
bc\ bc \ \bar{b}c
\]
During the trace abstraction step, the first two signal events $bc$ can be mapped to $\{e_1, e_2\}$ since $a$ is not observable, and the last one $b'c$ is mapped to $\{e_3\}$.  Therefore, this signal trace is abstracted to four flow traces, $\{e_1, e_2\} \times \{e_1, e_2\} \times \{e_3\}$.

Next, we consider the case where a flow event is mapped from a sequence of signal events.  Now suppose that two other flow events are implemented by sequences of signal events as defined in the list below.
\[
\begin{array}{cl}
e_4: & abc\ \bar{a}bc\\
e_5: & abc\ abc\ abc\ \bar{a}bc\\
\end{array}
\] 
Given an observed trace of the same observability shown below
\[
bc\ bc\ bc \ bc,
\]
it is abstracted to the following flow traces.
\[
e_4e_4,\; \textvisiblespace e_4 \textvisiblespace,\;  e_4 \textvisiblespace \textvisiblespace,\;  \textvisiblespace \textvisiblespace e_4, \; e_5
\] 
where $\textvisiblespace$ denotes signal events that are not mapped to any flow events.  Note that the above abstraction leads to three distinct flow traces as the middle three correspond to the same trace of flow events.

From the above discussions, it can be seen that a signal trace of partial observability is generally mapped to a set of flow traces.  As the the number of signals available for observation becomes smaller, the number of flow event traces corresponding to a signal trace as a result of the abstraction can be enormous.  This phenomenon can increase the complexity of the trace interpretation as it can cause the number of possible flow execution scenarios generated during the analysis to explode.   Possible solutions to address this issue include better trace signal selection for observation and assistance from debuggers' insights.  Trace signal selection itself is an important and difficult subject, and a detailed discussion of it is out of scope of this paper.   Next, we briefly describe how the debuggers' insights of a system's architecture can help to address the complexity issue in the trace analysis.

\subsection{Inputs from Validators}

During the trace interpretation, the number of intermediate flow execution scenarios may become too large due to ambiguous interpretation from signal events to flow events or from flow events to flow execution scenarios.  The explosion of the intermediate results can significantly slow down the performance of the trace analysis.  To address this problem, the debuggers can use their insights and understanding of the SUD to trim the total number of possible system executions derived from the trace analysis.  When a SUD is validated, a debugger is assumed to have deep knowledge about the system's architecture and microarchitecture, and how the test environment affects the system executions.  For instance, he/she may have knowledge that in a test environment, the maximal number of instances of a flow can be activated by the SUD, or a flow can be activated after certain other flows have terminated, etc.  These debugging insights can be encoded as constraints into the trace analysis method, which can then be used to eliminate a large number of flow execution scenarios that violate these constraints during the trace interpretation step.

@@

Alternatively, if all derived execution scenarios seem to be plausible, the implication that a debugger may draw from this result is that the failure may be independent of the flows being observed.  Therefore, the testing environment can be adjusted in order for a different part or different behavior of the SUD to be observed.  This idea, closely related to trace signal selection, is critical for post-silicon validation, and a detailed discussion can only be presented in a separate paper.

@@

The proposed method is demonstrated on a transaction level model of a simple SoC design with CPU core, memory, and a peripheral device connected by a highly concurrent interconnect in SystemC.  Since the proposed method is communication centric, the detailed computations of these blocks are not modeled.  Instead, the modeling is focused on how they participate in system level protocols.  This simple case study shows that the proposed method is capable of providing useful information to help reveal and locate various bugs in the SystemC model, thus making the debugging more efficient.  Once a bug-free model is available, the trace analysis method can correctly derive the instances of system flows activated by various components in the design from the observed traces.  More details about the design model and system flows implemented by the design model will be provided in a supplement document stored online.

@@

For this model, we consider the flow specifications describing the cache coherence protocols supported in GEM5 that is used to build the model in Fig~\ref{SoC}.  These flow specifications describe data/instruction read operations and data write operations initiated from CPUs.  Three such flows describe the cache coherent protocols for each CPU.  As there are two CPUs in this model, there are actually six such flows considered for this model.  These flow specifications are derived from the cache coherence protocols supported in GEM5\footnote{The GEM5 cache coherence protocols can be found at \url{http://www.gem5.org/docs/html/gem5MemorySystem.html}.}.  More detailed information about the flow specifications used in this case study will be provided as a supplement document if this paper is accepted for publication.   
%More detailed information about these protocols and the flow specifications can be found online\footnote{\url{https://github.com/cao2/paper/blob/master/Flow\%20Specification.pdf}} to meet the page limit.

@@

Two simple programs are written, one for each CPU.  These simple programs read numbers from a file, perform some operations on these numbers, and store the results back to the file.  How GEM5 supports shared memory multi-threaded program execution is unclear.  Therefore, there is no data shared in both caches in this test.  Furthermore, GEM5 does not support true concurrency.  When there are two programs running on the CPUs, GEM5 alternates the executions between the two CPUs.  To simulate asynchronous concurrency with the interleaving semantics, those two simple programs are instrumented with pseudo-blocking commands, one placed before each statement.  A pseudo blocking command includes a random number generator that returns either $0$ or $1$ and a loop that only exits when the returned random number is $0$.

>>>>

\begin{abstract}
We consider the problem of reconstructing system-level
behavior of an SoC design from a partially observed signal
trace.  Solving this problem is a critical activity in
post-silicon validation, and currently depends primarily on
human creativity and insights.  In this paper, we provide an
algorithm to automatically infer system-level transactions
from incomplete, ambiguous, and noisy trace data.  We
demonstrate the approach on a multicore virtual platoform
developed within the GEM5 environment.
\end{abstract}

@@

In this paper, we consider the problem of reconstructing
protocol-level behavior from silicon traces in SoC designs.
Given a collection of system-level communication protocols
and a trace of (partially observed) hardware signals, our
approach infers, with a certain measure of confidence, the
protocol instances (and their interleavings) being exercised
by the trace.  The approach is based on a formalization of
system-level transactions via labeled Petri-Nets, which are
capable of describing sequencing, concurrency, and choices
over system events.  We develop algorithms to infer
system-level transactions from traces with missing, noisy,
and ambiguous signal values.  We demonstrate our approach on
a multicore virtual platform constructed within the GEM5
environment~\cite{Binkert2011}.

@@

An SoC design involves integration of a number of IPs that
communicate through complex protocols.  Such system-level
protocols are typically specified in architecture documents
as message flow diagrams.  For this paper, we use the words
``protocol'' and ``flow'' interchangeably.
Fig~\ref{flow-spec-ex}(a) shows one diagram for a protocol
to authenticate and load a firmware during system boot for
firmware upgrade.  During validation, the system under debug
(SUD) exercises some complex system-level use-case which
involves interleaved execution of possibly a large number of
such flows.  A trace of a small number of hardware signals
is then shipped off-chip analysis.  The off-chip analysis
includes two broad phases: (1)~trace abstraction, and
(2)~trace interpretation.  Trace abstraction maps the
hardware trace into higher-level architectural constructs,
\eg, messages, operations, etc.: a message such as {\tt
  Authorization request} may be implemented in hardware
through a Boolean or temporal combination of specific
hardware signals in the NoC fabric between {\tt Device} and
{\tt CE}, \eg, as a sequence containing a header, a specific
value of a sequence of data words, etc.  We will refer to
such architectural constructs as {\em protocol events} or
{\em flow events}.  Note that due to limited observability,
it may not be possible to map a given set of (observed)
hardware signals uniquely to a flow event.  Finally, the
trace may a result from several instances of the same
protocol executing concurrently, \eg, a firmware
authentication protocol may be invoked when another instance
of the protocol has not completed.

@@

To illustrate the basic idea, consider the system flow in
Fig.~\ref{flow-spec-ex}(b), which we will call $F_1$.
Suppose that the following flow trace is abstracted from an
observed signal trace of the system.
\[
	t_1\;t_2\;t_1\;t_2\;t_3\;t_3\;t_4\;t_5\;t_5\;t_4\ldots
\]  
Here transition names in the LPN are used to represent the 
flow events in the trace.  The first four events results in
the following flow execution scenario
\[
	\{(F_{1,1}, \{p_3\}),~(F_{1,2}, \{p_3\})\}.
\]
For the first event $t_3$, it results in two execution scenarios 
below depending on which flow instance emits $t_3$.
\[
\begin{array}{l}
	\{(F_{1,1}, \{p_4\}),~(F_{1,2}, \{p_3\})\} \\
	\{(F_{1,1}, \{p_3\}),~(F_{1,2}, \{p_4\})\}.
\end{array}
\]
After handing the next event $t_3$, the above two execution scenarios
are reduced to the one as shown below.
\[
	\{(F_{1,1}, \{p_4\}),~(F_{1,2}, \{p_4\})\}.
\]
Using Algorithm~\ref{algo:compliance} to handle the
remaining four events, the following execution scenario is
derived.
\[
	\{(F_{1,1}, \{p_5, p_6\}),~(F_{1,2}, \{p_5, p_6\})\}
\]
%% Given a trace of flow events $\rho = e_1e_2\ldots e_n$, the
%% trace interpretation algorithm starts with an empty set of
%% of flow execution scenario $\mathit{Scen} = \emptyset$.
%% Then, for each $e_h$ where $1 \leq h \leq n$ starting $h=1$,
%% and for each $\mathit{scen} \in \mathit{Scen}$, the
%% following two steps are performed.
%% \begin{description}
%% \item[Step 1]~~For each $(F_{i,j}, s_{i,j}) \in
%%   \mathit{scen}$, if $\mathit{accept}(F_{i,j}, s_{i,j}, e_h)
%%   = (F_{i,j}, s'_{i,j})$, create a new scenario
%%   $\mathit{scen}' = (\mathit{scen} - (F_{i,j}, s_{i,j}))
%%   \cup (F_{i,j}, s'_{i,j})$, which is added into
%%   $\mathit{Scen}'$.

@@

@@

Hereafter, the term {\em flow traces} is used to refer to
traces of flow events.  Consider the following example for
the first case.  Suppose that there are three flow events:
$e_1$, $e_2$, and $e_3$, which are implemented in hardware
by the signal events shown in the list below.  We use
Boolean expressions to represent signal events for the
discussion.
\[
\begin{array}{cl}
e_1: & abc\\
e_2: & \bar{a}bc\\
e_3: & a\bar{b}c
\end{array}
\] 
Suppose that only signals $b$ and $c$ are observable, and we
obtain the following trace:
\[
bc\ bc \ \bar{b}c
\]
During trace abstraction, the first two signal events $bc$
can be mapped to $\{e_1, e_2\}$ since $a$ is not observable,
and the last one $b'c$ is mapped to $\{e_3\}$.  Therefore,
this signal trace is abstracted to four flow traces, $\{e_1,
e_2\} \times \{e_1, e_2\} \times \{e_3\}$.

Next, we consider the case where a flow event is mapped from
a sequence of signal events.  Now suppose that two other
flow events are implemented by sequences of signal events as
defined in the list below.
\[
\begin{array}{cl}
e_4: & abc\ \bar{a}bc\\
e_5: & abc\ abc\ abc\ \bar{a}bc\\
\end{array}
\] 
Given an observed trace of the same observability shown below
\[
bc\ bc\ bc \ bc,
\]
it is abstracted to the following flow traces.
\[
e_4e_4,\; \textvisiblespace e_4 \textvisiblespace,\;  e_4 \textvisiblespace \textvisiblespace,\;  \textvisiblespace \textvisiblespace e_4, \; e_5
\]

where $\textvisiblespace$ denotes signal events that are not
mapped to any flow events.  Note that the above abstraction
leads to three distinct flow traces as the middle three
correspond to the same trace of flow events.

It is clear from above that a partial trace is viewed as a
set of flow traces, and Algorithm~\ref{algo:compliance} can
be suitably extended to work with flow traces to obtain the
set of candidate flows.  However, applicability of the
algorithm in practice can be gated because the number of
potential flow execution scenarios generated under partial
observability may be enormous.  Note that this is not a
limitation of the algorithm; if the observability of
critical events is poor there simply {\em are} too many flow
execution scenarios compliant with the observed trace.
Nevertheless, we need to address the issue to make trace
interpretation (whether automatic or not) practicable.
There are two potential approaches: (1)~better selection of
post-silicon trace observability, and (2)~use of system
insights during validation.  Trace signal selection itself
is an important and orthogonal topic~\cite{nicolici,basu},
and a detailed discussion of it is out of scope of this
paper.  However, we briefly describe how the debuggers'
insights of a system's architecture can help to address the
complexity issue in the trace trace interpretation.

\subsection{Interactive Trace Interpretation}

Post-silicon validation is performed by debuggers with deep
knowledge about the system's architecture and
microarchitecture, and the test environment.  Two key
insights are (1) the maximal number of instances of a flow
activated in the test environment, and (2)~the mutual
relationship between two flows.  For example, the test
environment may not permit multiple instances of firmware
authentication to operate concurrently, or a flow involving
audio and Web browsing to initiate until the flows
participating in boot are completed.  Our framework permits
incorporating such insights as constraints in trace
analysis; flow execution scenarios that violate these
constraints are ignored.  Apart from the explosion problem,
this permits the debugger to quickly filter out
uninteresting combination of flows and focus on interesting
interleavings.

@@

%% Alternatively, if all derived execution scenarios seem to be
%% plausible, the implication that a debugger may draw from
%% this result is that the failure may be independent of the
%% flows being observed.  Therefore, the testing environment
%% can be adjusted in order for a different part or different
%% behavior of the SUD to be observed.  This idea, closely
%% related to trace signal selection, is critical for
%% post-silicon validation, and a detailed discussion can only
%% be presented in a separate paper.

@@

%% The proposed method is demonstrated on a transaction level
%% model of a simple SoC design with CPU core, memory, and a
%% peripheral device connected by a highly concurrent
%% interconnect in SystemC.  Since the proposed method is
%% communication centric, the detailed computations of these
%% blocks are not modeled.  Instead, the modeling is focused on
%% how they participate in system level protocols.  This simple
%% case study shows that the proposed method is capable of
%% providing useful information to help reveal and locate
%% various bugs in the SystemC model, thus making the debugging
%% more efficient.  Once a bug-free model is available, the
%% trace analysis method can correctly derive the instances of
%% system flows activated by various components in the design
%% from the observed traces.  More details about the design
%% model and system flows implemented by the design model will
%% be provided in a supplement document stored online.

@@

For this model, we consider the flow specifications
describing the cache coherence protocols supported in GEM5
that is used to build the model in Fig~\ref{SoC}.  These
flow specifications describe data/instruction read
operations and data write operations initiated from CPUs.
Three such flows describe the cache coherent protocols for
each CPU.  As there are two CPUs in this model, there are
actually six such flows considered for this model.  These
flow specifications are derived from the cache coherence
protocols supported in GEM5.\footnote{The GEM5 cache
  coherence protocols can be found at
  \url{http://www.gem5.org/docs/html/gem5MemorySystem.html}.}

%% More detailed information about the flow specifications used in this case study will be provided as a supplement document if this paper is accepted for publication.   
%More detailed information about these protocols and the flow specifications can be found online\footnote{\url{https://github.com/cao2/paper/blob/master/Flow\%20Specification.pdf}} to meet the page limit.

@@

We wrote two simple concurrent programs --- one for each CPU
--- to exercise the flows.  They read numbers from a file,
perform some operations on these numbers, and store the
results back to the file.  How GEM5 supports shared memory
multi-threaded program execution is unclear.  Therefore,
there is no data shared in both caches in this test.
Furthermore, GEM5 does not support true concurrency.  When
there are two programs running on the CPUs, GEM5 alternates
the executions between the two CPUs.  To simulate
asynchronous concurrency with the interleaving semantics,
those two simple programs are instrumented with
pseudo-blocking commands, one placed before each statement.
A pseudo blocking command includes a random number generator
that returns either $0$ or $1$ and a loop that only exits
when the returned random number is $0$.
